Towards Practical First-Order Model Counting

### 2024-07-15 doc/paper/paper.tex
Some editing

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{EQ}$}
\newcommand{\Cranetwo}{\textsc{Crane2}}
\newcommand{\Cranebfs}{\textsc{Crane2-BFS}}
\newcommand{\Cranegreedy}{\textsc{Crane2-Greedy}}

@@

\SetKwFunction{CompileWithBaseCases}{CompileWithBaseCases}
\SetKwFunction{Compile}{Compile}
\SetKwFunction{Propagate}{Propagate}
\SetKwFunction{FindBaseCases}{FindBaseCases}
\SetKwFunction{Simplify}{Simplify}

@@

% TODO: No proper runtime consideration (other than the termination). (address
% this criticism in some way)

@@

% TODO: Updates to the algorithm
% 1) replace x_i-c_i on Line 7 of Algorithm 2 with a generalisation of x_i that can be an arbitrary summation/subtraction of domain sizes
% 2) a function call with a constant should always mean that that function should have a base case, not just when it's in the same function

% TODO: consider restructuring the paper to make it easier to understand

\section{Introduction}
% TODO: improvements to the introduction (must):
% * make it extra clear what the gap is that this work is trying to fill
% * mention Crane earlier and describe its importance
% * the last two paragraphs should make more of a BOOM (e.g., include the over 9k claim)
% * (here and elsewhere) clarify the distinction between the weighted and unweighted variants of the problem in the final version of the paper. We focus on the unweighted variant in our implementation and experiments, however, Crane2 can handle WFOMC instances as well simply by switching the main data type from integers to rational numbers. Maybe add some weighted experiments as well.
% * (must) more related work, situating my contributions in the broader context
% * (concerning related work) the topic is intimately related to "graded logics", largely investigated in DL, mu-calculus and temporal logics

@@

There are various WFOMC algorithms with different underlying principles. Perhaps
the most prominent class of WFOMC algorithms is based on \emph{first-order
  knowledge compilation}. In this approach, the formula is compiled into a
representation (such as a circuit or graph) by iteratively applying
\emph{compilation rules}. This representation can then be used to compute the
WFOMC for any combination of domain sizes (or weights). Algorithms in this class
include \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its recent
extension \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. The former compiles
formulas into circuits, while the latter compiles them first to
\emph{first-order computational graphs} (FCGs) and then to (algebraic)
equations. Another WFOMC algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, is based on cell enumeration.
Other algorithms utilise local search~\cite{DBLP:journals/pvldb/NiuRDS11},
junction trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.
% TODO: FCGs were never defined. Maybe avoid referring to them altogether?
% Smoothing nodes are also undefined.

@@

% TODO: first sentence: elaborate a bit more for a general audience
The recently proposed \textsc{Crane} algorithm, while capable of handling
formulas beyond the reach of \textsc{ForcLift}, was incomplete as it could only
construct function definitions, which would then need to be evaluated to compute
the WFOMC.\@ Additionally, recursive functions were presented without base
cases. In this work, we introduce \Cranetwo{}, an extension of \textsc{Crane}
that addresses these two weaknesses.
% TODO: We called the previous version of Crane "incomplete" because it cannot
% be used as a black box that takes Markov logic networks (or weighted formulas
% in some other format) and outputs numbers (i.e., model counts or
% probabilities).

@@

    \node[draw,fit={(compilewithbasecases) (compilation) (crane) (findbasecases) (propagate)},inner ysep=7pt,yshift=5pt] {};
    \node at (0.2, 0.5) {\Cranetwo};

@@

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{The outline of using \Cranetwo{} to compute the model count of a formula $\phi$. First, the formula is compiled into a set of equations, which are then used to create a C++ program. This program can be executed with different command line arguments to calculate the model count of $\phi$ for different domain sizes. To accomplish this, the \CompileWithBaseCases function employs several components:
    \begin{enumerate*}[label=(\roman*)]
      \item the knowledge compilation algorithm of \textsc{Crane}, referred to as \Compile,
      \item a procedure called \FindBaseCases, which identifies a sufficient set of base cases,
      \item a procedure called \Propagate, which constructs a formula corresponding to a given base case, and
      \item algebraic simplification techniques (denoted as \Simplify).
    \end{enumerate*}
  }\label{fig:overview}
\end{figure*}

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas corresponding to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} explains post-processing techniques for the formulas
from \cref{step2} above required to preserve the correct model count. Next,
\cref{sec:cpp} elucidates how function definitions encoding a solution to a
WFOMC problem are compiled into C++ programs.\footnote{The possibility of
  compiling WFOMC solutions to C++ programs has been previously
  considered~\cite{DBLP:conf/kr/KazemiP16}. However, the extent of formulas that
  could be handled was limited.} Finally, \cref{sec:experiments} presents two
experiments comparing \Cranetwo{} with other state-of-the-art WFOMC algorithms.

@@

In this section, we will review the basic concepts of first-order logic as they
are used in first-order knowledge compilation algorithms. There are two key
differences between the logic used within such algorithms and the logic
supported as input. First, Skolemization~\cite{DBLP:conf/kr/BroeckMD14} is
employed to eliminate existential quantifiers by introducing additional
predicates. Second, the input formula is rewritten as a conjunction of clauses,
each of which is in \emph{prenex normal form}~\cite{hinman2018fundamentals}.
% TODO: explain that "Skolemization" here is NOT standard Skolemization of
% first-order logic, which would replace existential quantifiers by function
% terms.

% TODO: reorder the definitions in the opposite direction
A \emph{formula} is a conjunction of clauses. A \emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\text{.}\forall x_{2} \in \Delta_{2}\dots\text{
}\forall x_{n} \in \Delta_{n}\text{.}\phi(x_{1}, x_{2}, \dots, x_{n})$, where
$\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). We say that a clause is a
\emph{(positive) unit clause} if
\begin{enumerate*}[label=(\roman*)]
  \item there is only one literal with a predicate, and
  \item it is a positive literal.
\end{enumerate*}
A \emph{literal} is either an atom (i.e., a \emph{positive} literal) or its
negation (i.e., a \emph{negative} literal). An \emph{atom} is either
\begin{enumerate*}[label=(\roman*)]
  \item $P(t_{1}, \dots, t_{m})$ for some predicate $P$ and terms
  $t_{1}, \dots, t_{m}$ (written as $P(\mathbf{t})$ for short) or
  \item $x=y$ for some terms $x$ and $y$.
\end{enumerate*}
An atom is \emph{ground} if it contains no variables, i.e., only constants. The
\emph{arity} of a predicate is the number of arguments it takes, i.e., $m$ in
the case of predicate $P$ mentioned above. When we want to denote a predicate
together with its arity, we write $P/m$. A \emph{term} is either a variable or a
constant. Throughout the paper, we will use set-theoretic notation, interpreting
a formula as a set of clauses and a clause as a set of literals. Moreover, for
readability, clauses written on separate lines are implicitly conjoined.

@@

% TODO: mention the Unique Name Assumption and that, hence, s = t holds iff s
% and t are the same constant.

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{Comparison of three logics used in WFOMC based on the following aspects:
    \begin{enumerate*}[label=(\roman*)]
      \item number of sorts,
      \item support for constants,
      \item maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

% TODO:
% 1. Rename \UFO{} to what that new paper uses
% 2. Use FastWFOMC.jl instead, removing the mention of 'privately-obtained'
% 3. use symbols instead of numbers for the footnotes
In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in WFOMC:
\begin{enumerate*}[label=(\roman*)]
  \item \FO{} is the input format for
  \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
  its extensions
  \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
  \Cranetwo{};
  \item \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
  related
  methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22};
  \item \UFO{} is the input format supported by a private version of
  \textsc{FastWFOMC} obtained directly from the authors.
\end{enumerate*}
Note that the publicly available
version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
of \textsc{FastWFOMC} does not support any cardinality constraints. The notation
we use to refer to each logic is standard in the case of \Ctwo{}, new in the
case of \UFO{}, and redefined to be more specific in the case of \FO{}. All
three logics are function-free, and domains are always assumed to be finite.

In \FO{}, each term is assigned to a \emph{sort}, and each predicate $P/n$ is
assigned to a sequence of $n$ sorts. Each sort has its corresponding domain.
Most of these assignments to sorts are typically left implicit and can be
reconstructed from the quantifiers. For example, $\forall x,y \in \Delta\text{.
}P(x, y)$ implies that variables $x$ and $y$ have the same sort. On the other
hand, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{. } P(x, y)$
implies that $x$ and $y$ have different sorts, and it would be improper to
write, for example, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{.
} P(x, y) \lor x = y$. \FO{} is also the only logic to support constants,
formulas with more than two variables, and the equality predicate.
% TODO: the discussion about sorts isn't featured in the rest of the paper

@@

\Ctwo{} and \UFO{} are single-sorted (i.e., all variables are quantified over
the same domain) and only include formulas with at most two variables and no
constants. Unlike \UFO{}, \Ctwo{} supports existential quantifiers, including
\emph{counting quantifiers}. For example, $\exists^{=1} x\text{. }\phi(x)$ means
that there exists \emph{exactly one} $x$ such that $\phi(x)$, and
$\exists^{\le 2} x\text{. }\phi(x)$ means that there exist \emph{at most two}
such $x$. The advantage \UFO{} brings over \Ctwo{} is the support for
\emph{(equality) cardinality constraints}. For example, $|P| = 3$ constrains all
models to have \emph{precisely three positive literals with the predicate $P$}.

% TODO: The distinctness of domains is important in two ways. First, in terms of
% expressiveness, writing clause such as "\forall x, P(x, x)" is valid if
% predicate P is defined over two copies of the same domain and invalid
% otherwise. Second, having more distinct domains makes the problem more
% decomposable for the knowledge compilation algorithm. With more distinct
% domains, it can make assumptions or deductions about, e.g., the first domain
% of predicate P without worrying how (or if) they apply to the second domain.
\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains (which
  may not be distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets, satisfying the following conditions:
  \begin{enumerate*}[label=(\roman*)]
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{enumerate*}
  Then a \emph{structure} of $\phi$ (with respect to $\sigma$) is a set $M$ of
  ground literals defined by adding either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it satisfies $\phi$.
\end{definition}
% TODO: adding to what

% TODO: “Weights in Q” undefined, C2 undefined (and also far from well explained
% before), endofunction undefined (also endomorphisms missing), weight function
% undefined, weighted model count imprecise (above it’s called weighted
% first-order model count/wfmoc),
\begin{definition}[WFOMC~\cite{DBLP:conf/ijcai/BroeckTMDR11}]
  Continuing from \cref{def:model}, for each predicate $P$ in $\phi$, let
  $w^{+}(P)$ and $w^{-}(P)$ be its \emph{weights} in $\mathbb{Q}$. Unless
  specified otherwise, we assume all weights are equal to 1. The
  \emph{(symmetric) weighted first-order model count} (WFOMC) of $\phi$ (with
  respect to $\sigma$, $w^{+}$, and $w^{-}$) is:
  \[
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \]
  where the sum is over all models of $\phi$.
\end{definition}

\begin{example}[Counting functions]\label{example:functions}
  To define predicate $P$ as an endofunction on $\Delta$, in \Ctwo{} one would
  write $\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{.
  }P(x, y)$. In \UFO{}, the same could be written as
  \begin{equation}\label[formula]{eq:functions1}
    \begin{gathered}
      \forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)\\
      |P| = |\Delta|,
    \end{gathered}
  \end{equation}
  where $w^{-}(S) = -1$. Although \cref{eq:functions1} has more models compared
  to its counterpart in \Ctwo{}, the weight function makes the \emph{weighted}
  model count of \cref{eq:functions1} equal to the number of endofunctions on
  $\Delta$.

  % TODO: Yes, as mentioned at the end of Section 2.1, clauses c and d written
  % on separate lines are to be interpreted as the conjunction of c and d (maybe
  % move that sentence here?).
  Equivalently, in \FO{} we would write
  \begin{equation}\label[formula]{eq:fo}
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z.
    \end{gathered}
  \end{equation}
  The first clause asserts that each $x$ must have at least one corresponding
  $y$, while the second statement adds the condition that if $x$ is mapped to
  both $y$ and $z$, then $y$ must equal $z$. It is important to note that
  \cref{eq:fo} is written with two domains instead of just one. However, we can
  still determine the number of endofunctions by assuming that the sizes of
  $\Gamma$ and $\Delta$ are equal. This formulation, as observed by
  \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}, can
  prove beneficial in enabling first-order knowledge compilation algorithms to
  find efficient solutions.
\end{example}

@@

\section{Completing the Definitions of Recursive Functions}\label{sec:main}
% TODO: As mentioned in the caption of Figure 1, "Compile" refers to the
% knowledge compilation algorithm of Dilkas and Belle (2023). Its three return
% values are explained at the beginning of page 4. We do not elaborate on the
% details of "Simplify" as it primarily consists of elementary algebraic
% simplifications such as "1x = x" and "0x = 0". However, we highlight one
% non-trivial aspect of "Simplify" right before Section 3.1. We will improve the
% structure of Section 3 to make this information more readily available. Maybe
% write Crane instead of Compile?

>>>>

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{CC}$}
\newcommand{\Cranetwo}{\textsc{Crane2}}
\newcommand{\Cranebfs}{\textsc{Crane2-BFS}}
\newcommand{\Cranegreedy}{\textsc{Crane2-Greedy}}

@@

\SetKwFunction{CompileWithBaseCases}{CompileWithBaseCases}
\SetKwFunction{Compile}{{\normalfont \textsc{Crane}}}
\SetKwFunction{Propagate}{Propagate}
\SetKwFunction{FindBaseCases}{FindBaseCases}
\SetKwFunction{Simplify}{Simplify}

@@

@@

% TODO: consider restructuring the paper (how?)

\section{Introduction}
% TODO: make it extra clear what the gap is that this work is trying to fill

% TODO: mention Crane earlier and describe its importance

% TODO: the last two paragraphs should make more of a BOOM (e.g., include the
% over 9k claim)

% TODO (here and elsewhere): clarify the distinction between the weighted and
% unweighted variants of the problem in the final version of the paper. We focus
% on the unweighted variant in our implementation and experiments, however,
% Crane2 can handle WFOMC instances as well simply by switching the main data
% type from integers to rational numbers. Maybe add some weighted experiments as
% well.

% TODO: (must) more related work, situating my contributions in the broader
% context

% TODO: (concerning related work) the topic is intimately related to "graded
% logics", largely investigated in DL, mu-calculus and temporal logics

@@

There are various WFOMC algorithms with different underlying principles. Perhaps
the most prominent class of WFOMC algorithms is based on \emph{first-order
  knowledge compilation} (FOKC). In this approach, the formula is compiled into
a representation (such as a circuit or graph) by iteratively applying
\emph{compilation rules}. This representation can then be used to compute the
WFOMC for any combination of domain sizes (or weights). Algorithms in this class
include \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its recent
extension \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. The former compiles
formulas into circuits, while the latter compiles them first to graphs and then
to (algebraic) equations. Another WFOMC algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, is based on cell enumeration.
Other algorithms utilise local search~\cite{DBLP:journals/pvldb/NiuRDS11},
junction trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.

@@

% TODO: introduce MLNs somewhere? maybe use an acronym?
The recently proposed \textsc{Crane} algorithm, while capable of handling
formulas beyond the reach of \textsc{ForcLift}, was `incomplete' as it could
only construct function definitions, which would then need to be evaluated to
compute the WFOMC.\@ In other words, it could not be used as a black box that
takes Markov logic networks (or weighted formulas in some other format) and
outputs numbers (i.e., model counts or probabilities). Additionally, recursive
functions were presented without base cases. In this work, we introduce
\Cranetwo{}, an extension of \textsc{Crane} that addresses these two weaknesses.

@@

    \node[draw,fit={(compilewithbasecases) (compilation) (crane) (findbasecases) (propagate)},inner ysep=7pt,yshift=5pt] {};
    \node at (0.6, 0.5) {\Cranetwo};

@@

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{The outline of using \Cranetwo{} to compute the model count of a formula $\phi$. First, the formula is compiled into a set of equations, which are then used to create a C++ program. This program can be executed with different command line arguments to calculate the model count of $\phi$ for different domain sizes. To accomplish this, the \CompileWithBaseCases function employs several components:
    \begin{enumerate*}[label=(\roman*)]
      \item the FOKC algorithm of \textsc{Crane},
      \item a procedure called \FindBaseCases, which identifies a sufficient set of base cases,
      \item a procedure called \Propagate, which constructs a formula corresponding to a given base case, and
      \item algebraic simplification techniques (denoted as \Simplify).
    \end{enumerate*}
  }\label{fig:overview}
\end{figure*}

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas corresponding to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} explains post-processing techniques for the formulas
from \cref{step2} above required to preserve the correct model count. Next,
\cref{sec:cpp} elucidates how function definitions encoding a solution to a
WFOMC problem are compiled into C++ programs. Note that a solution to WFOMC that
uses compilation to C++ has been considered
before~\cite{DBLP:conf/kr/KazemiP16}, however, the extent of formulas that could
be handled was limited. Finally, \cref{sec:experiments} presents experiments
results comparing \Cranetwo{} with other state-of-the-art WFOMC algorithms.

@@

In this section, we review the basic concepts of first-order logic as they are
used in FOKC algorithms. There are two key differences between the logic used by
such algorithms and the logic supported as input. First,
Skolemization~\cite{DBLP:conf/kr/BroeckMD14} is employed to eliminate
existential quantifiers by introducing additional predicates. Note that
Skolemization here is different from the standard Skolemization procedure that
introduces function symbols~\cite{DBLP:books/daglib/0030198}. Second, the input
formula is rewritten as a conjunction of clauses, each of which is in
\emph{prenex normal form}~\cite{hinman2018fundamentals}.

A \emph{term} is either a variable or a constant. An \emph{atom} is either
\begin{enumerate*}[label=(\roman*)]
  \item $P(t_{1}, \dots, t_{m})$ for some predicate $P$ and terms
  $t_{1}, \dots, t_{m}$ (written as $P(\mathbf{t})$ for short) or
  \item $x=y$ for some terms $x$ and $y$.
\end{enumerate*}
The \emph{arity} of a predicate is the number of arguments it takes, i.e., $m$
in the case of predicate $P$ above. We write $P/m$ to denote a predicate
together with its arity. A \emph{literal} is either an atom (i.e., a
\emph{positive} literal) or its negation (i.e., a \emph{negative} literal). An
atom is \emph{ground} if it contains no variables, i.e., only constants. A
\emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\text{.}\forall x_{2} \in \Delta_{2}\dots\text{}\forall x_{n} \in \Delta_{n}\text{.}\phi(x_{1}, x_{2}, \dots, x_{n})$,
where $\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). We say that a clause is a
\emph{(positive) unit clause} if
\begin{enumerate*}[label=(\roman*)]
  \item there is only one literal with a predicate, and
  \item it is a positive literal.
\end{enumerate*}
Finally, a \emph{formula} is a conjunction of clauses. Throughout the paper, we
use set-theoretic notation, interpreting a formula as a set of clauses and a
clause as a set of literals.

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in WFOMC based on the following aspects:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

% TODO: Use FastWFOMC.jl instead, removing the mention of 'privately-obtained'

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in WFOMC:
\begin{enumerate*}[label=(\roman*)]
  \item \FO{} is the input format for
  \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
  its extensions
  \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
  \Cranetwo{};
  \item \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
  related
  methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22};
  \item \UFO{} is the input format supported by a private version of
  \textsc{FastWFOMC} obtained directly from the authors.
\end{enumerate*}
Note that the publicly available
version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
of \textsc{FastWFOMC} does not support any cardinality constraints. The notation
we use to refer to each logic is standard in the case of \Ctwo{} and redefined
to be more specific in the case of the other two logics. All three logics are
function-free, and domains are always assumed to be finite. As usual, we
presuppose the \emph{unique name assumption}, which states that two constants
are equal if and only if they are the same constant~\cite{DBLP:books/aw/RN2020}.

\renewcommand*{\thefootnote}{\arabic{footnote}}

In \FO{}, each term is assigned to a \emph{sort}, and each predicate $P/n$ is
assigned to a sequence of $n$ sorts. Each sort has its corresponding domain.
These assignments to sorts are typically left implicit and can be reconstructed
from the quantifiers. For example, $\forall x,y \in \Delta\text{. }P(x, y)$
implies that variables $x$ and $y$ have the same sort. On the other hand,
$\forall x \in \Delta\text{. }\forall y \in \Gamma\text{. } P(x, y)$ implies
that $x$ and $y$ have different sorts, and it would be improper to write, for
example, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{.
} P(x, y) \lor x = y$. \FO{} is also the only logic to support constants,
formulas with more than two variables, and the equality predicate. While we do
not explicitly refer to sorts in subsequent sections of this paper, the
many-sorted nature of \FO{} is paramount to the algorithms presented therein.

@@

Compared to \FO{}, \Ctwo{} and \UFO{} lack support for:
\begin{enumerate*}[label=(\roman*)]
  \item constants,
  \item the equality predicate,
  \item multiple domains, and
  \item formulas with more than two variables.
\end{enumerate*}
The advantage that \Ctwo{} brings over \FO{} is the inclusion of \emph{counting
  quantifiers}. That is, alongside $\forall$ and $\exists$, \Ctwo{} supports
$\exists^{=k}$, $\exists^{\le k}$, and $\exists^{\ge k}$ for any positive
integer $k$. For example, $\exists^{=1} x\text{. }\phi(x)$ means that there
exists \emph{exactly one} $x$ such that $\phi(x)$, and $\exists^{\le 2} x\text{.
}\phi(x)$ means that there exist \emph{at most two} such $x$. \UFO{}, on the
other hand, does not support any existential quantifiers but instead
incorporates \emph{(equality) cardinality constraints}. For example, $|P| = 3$
constrains all models to have \emph{precisely three positive literals with the
  predicate $P$}.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains (which
  may not be distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets, satisfying the following conditions:
  \begin{enumerate*}[label=(\roman*)]
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{enumerate*}
  (In practice, we typically only specify the size of each domain.) A
  \emph{structure} of $\phi$ (with respect to $\sigma$) is a set $M$ of ground
  literals defined by adding to $M$ either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it satisfies $\phi$.
\end{definition}

\begin{remark}
  The distinctness of domains is important in two ways. First, in terms of
  expressiveness, clauses such as $\forall x \in \Delta\text{. }P(x, x)$ are
  valid if predicate $P$ is defined over two copies of the same domain and
  invalid otherwise. Second, having more distinct domains makes the problem more
  decomposable for the FOKC algorithm. With more distinct domains, the algorithm
  can make assumptions or deductions about, e.g., the first domain of predicate
  $P$ without worrying how (or if) they apply to the second domain.
\end{remark}

\begin{definition}[WFOMC Instance]\label{def:instance}
  A \emph{WFOMC instance} comprises of:
  \begin{enumerate*}[label=(\roman*)]
    \item a formula $\phi$ in \FO{},
    \item two (rational) \emph{weights} $w^{+}(P)$ and $w^{-}(P)$ assigned to
    each predicate $P$ in $\phi$, and
    \item $\sigma$ as described in \cref{def:model}.
  \end{enumerate*}
  Unless specified otherwise, we assume all weights to be equal to 1.
\end{definition}

\begin{definition}[WFOMC~\cite{DBLP:conf/ijcai/BroeckTMDR11}]
  Given a WFOMC instance $(\phi, w^{+}, w^{-}, \sigma)$ as in
  \cref{def:instance}, the \emph{(symmetric) weighted first-order model count}
  (WFOMC) of $\phi$ (with respect to $\sigma$, $w^{+}$, and $w^{-}$) is
  \begin{equation}\label{eq:wfomc}
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \end{equation}
  where the sum is over all models of $\phi$.
\end{definition}

\begin{example}[Counting functions]\label{example:functions}
  To define predicate $P$ as a function from a domain $\Delta$ to itself, in
  \Ctwo{} one would write $\forall x \in \Delta\text{.
  }\exists^{=1} y \in \Delta\text{. }P(x, y)$. In \UFO{}, the same could be
  written as
  \begin{equation}\label[formula]{eq:functions1}
    \begin{gathered}
      (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)) \land{}\\
      (|P| = |\Delta|),
    \end{gathered}
  \end{equation}
  where $w^{-}(S) = -1$. Although \cref{eq:functions1} has more models compared
  to its counterpart in \Ctwo{}, the negative weight $w^{-}(S) = -1$ makes some
  of the terms in \cref{eq:wfomc} cancel out.

  Equivalently, in \FO{} we would write
  \begin{equation}\label[formula]{eq:fo}
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)) \land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z).
    \end{gathered}
  \end{equation}
  The first clause asserts that each $x$ must have at least one corresponding
  $y$, while the second statement adds the condition that if $x$ is mapped to
  both $y$ and $z$, then $y$ must equal $z$. It is important to note that
  \cref{eq:fo} is written with two domains instead of just one. However, we can
  still determine the correct number of functions by assuming that the sizes of
  $\Gamma$ and $\Delta$ are equal. This formulation, as observed by
  \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}, can
  prove beneficial in enabling FOKC algorithms to find efficient solutions.
\end{example}

@@

\section{Completing the Definitions of Functions}\label{sec:main}

\begin{algorithm}[t]
  \caption{\protect\CompileWithBaseCases{$\phi$}}\label{alg:compilewithbasecases}
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Compile{$\phi$}$\;\label{line:first}
  $\mathcal{E} \gets \Simplify{$\mathcal{E}$}$\;\label{line:second}
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{index $i$ such that $x_{i} \in \mathbb{N}_{0}$}{\label{line:loop}
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $\mathcal{E} \gets \mathcal{E} \cup \CompileWithBaseCases{$\psi$}$\;\label{line:final}
  }
\end{algorithm}
